---
title: "Regression Final Project"
output:
  html_document:
    df_print: paged
author: "Cheryl Allen-Munley and Varshini Yanamandra"
---

```{r}
library(readxl)	
library(car)
library(MASS)
library(TSA)
library(ggplot2)
library(tidyr)
library(broom)
```

```{r}
#Loading the political database
PolData = 	read_excel(file.choose())
PolData <- na.omit(PolData)
PolData				
```

```{r}
# converting string data into numbers
# 1 for Democrats and 2 for Republicans 
PolData$PresParty_label <- PolData$PresParty
PolData$PresParty <- as.factor(PolData$PresParty)
PolData$PresParty <- as.numeric(PolData$PresParty)
                                
PolData$SenParty_label <- PolData$SenParty
PolData$SenParty <- as.factor(PolData$SenParty)
PolData$SenParty <- as.numeric(PolData$SenParty)

PolData$HouseParty_label <- PolData$HouseParty
PolData$HouseParty <- as.factor(PolData$HouseParty)
PolData$HouseParty <- as.numeric(PolData$HouseParty)

PolData$TrifectaFlag_label <- PolData$TrifectaFlag
PolData$TrifectaFlag <- as.factor(PolData$TrifectaFlag)
PolData$TrifectaFlag <- as.numeric(PolData$TrifectaFlag)
PolData$TrifectaFlag <- replace(PolData$TrifectaFlag, PolData$TrifectaFlag == 2, 0)
```
```{r}
PolData$TrifectaFlag <- replace(PolData$TrifectaFlag, PolData$TrifectaFlag == 3, 2)
PolData
```
Creating preliminary models
```{r}
inflationModeli <- lm(MeanAnnualInflation ~ PresParty + SenParty + HouseParty + TrifectaFlag, data = PolData)
summary(inflationModeli)
```

```{r}
glance(inflationModeli)
```

```{r}
inflationModel <- lm(MeanAnnualInflation ~ PresParty + SenParty + HouseParty + TrifectaFlag - 1, data = PolData)
summary(inflationModel)
```
```{r}
minWageModeli <- lm(AdjMinWage ~ PresParty + SenParty + HouseParty + TrifectaFlag, data = PolData)
summary(minWageModeli)
```
```{r}
minWageModel <- lm(AdjMinWage ~ PresParty + SenParty + HouseParty + TrifectaFlag - 1, data = PolData)
summary(minWageModel)
```
```{r}
GDPModeli <- lm(GDP_AnnualGrowth ~ PresParty + SenParty + HouseParty + TrifectaFlag, data = PolData)
summary(GDPModeli)
```
```{r}
GDPModel <- lm(GDP_AnnualGrowth ~ PresParty + SenParty + HouseParty + TrifectaFlag - 1, data = PolData)
summary(GDPModel)
```
```{r}
unemploymentModeli <- lm(UnemploymentRateDecember ~ PresParty + SenParty + HouseParty + TrifectaFlag, data = PolData)
summary(unemploymentModeli)
```

```{r}
unemploymentModel <- lm(UnemploymentRateDecember ~ PresParty + SenParty + HouseParty + TrifectaFlag - 1, data = PolData)
summary(unemploymentModel)
```

Perform backward selection on each of the 4 response variable models. The no-intercept models are chosen as they have better R-squared values for every response variable.

```{r}
print("ADJUSTED R SQUARED VALUES")
print("Inflation model")
summary(inflationModeli)$adj.r.squared
print("Inflation model without intercept")
summary(inflationModel)$adj.r.squared

print("Minimum Wage model")
summary(minWageModeli)$adj.r.squared
print("Minimum Wage model without intercept")
summary(minWageModel)$adj.r.squared

print("GDP model")
summary(GDPModeli)$adj.r.squared
print("GDP model without intercept")
summary(GDPModel)$adj.r.squared

print("Unemployment model")
summary(unemploymentModeli)$adj.r.squared
print("Unemployment model without intercept")
summary(unemploymentModel)$adj.r.squared
```

Backward selection on models without intercept
```{r}
inflation.backward <- step(inflationModel, scope = list(lower=~1), direction = "backward", k=2, trace=FALSE)
summary(inflation.backward)
```

```{r}
minimumWage.backward <- step(minWageModel, scope = list(lower=~1), direction = "backward", k=2, trace=FALSE)
summary(minimumWage.backward)
```

```{r}
plot(minimumWage.backward)
```

```{r}
GDP.backward <- step(GDPModel, scope = list(lower=~1), direction = "backward", k=2, trace=FALSE)
summary(GDP.backward)
```

```{r}
unemployment.backward <- step(unemploymentModel, scope = list(lower=~1), direction = "backward", k=2, trace=FALSE)
summary(unemployment.backward)
```
Optimum: UnemploymentRateDecember ~ PresParty and HouseParty with no lag

```{r}
plot(unemployment.backward)
```

```{r}
boxplot(UnemploymentRateDecember~PolData$PresParty_label,data=PolData, xlab="Party", ylab = "Unemployment Rate" ,col=c("blue", "red"),main ="Unemployment Rate versus Presidential Party")
boxplot(UnemploymentRateDecember~HouseParty_label,data=PolData,xlab="Party", ylab = "Unemployment Rate",col=c("blue","red"),main =" Unemployment Rate versus House Party")
```

Explore the effect of lag on inflation
```{r}
# lag 1
PolData.inf <- PolData[, c(2, 3, 7, 12, 13)]
PolData.inf$MeanAnnualInflation <- c(NA, head(PolData.inf$MeanAnnualInflation, -1))
PolData.inf <- na.omit(PolData.inf)
PolData.infmodel1 <- lm(MeanAnnualInflation ~ PresParty + SenParty - 1, data = PolData.inf)
summary(PolData.infmodel1)
```

```{r}
# lag 2
PolData.inf$MeanAnnualInflation <- c(NA, head(PolData.inf$MeanAnnualInflation, -1))
PolData.inf <- na.omit(PolData.inf)
PolData.infmodel2 <- lm(MeanAnnualInflation ~ PresParty + SenParty - 1, data = PolData.inf)
summary(PolData.infmodel2)
```
Optimum: MeanAnnualInflation ~ PresParty and SenParty with lag 2

```{r}
plot(PolData.infmodel2)
```

```{r}
boxplot(MeanAnnualInflation~PresParty_label,data=PolData.inf, xlab="Party", ylab = "Inflation" ,col=c("blue", "red"),main ="Mean Annual Inflation versus Presidential Party")
boxplot(MeanAnnualInflation~SenParty_label,data=PolData.inf,xlab="Party", ylab = "Inflation", col=c("blue", "red"),main ="Mean Annual Inflation versus Senate Party")
```

```{r}
# lag 3
PolData.inf$MeanAnnualInflation <- c(NA, head(PolData.inf$MeanAnnualInflation, -1))
PolData.inf <- na.omit(PolData.inf)
PolData.infmodel3 <- lm(MeanAnnualInflation ~ PresParty + SenParty - 1, data = PolData.inf)
summary(PolData.infmodel3)
```

```{r}
# lag 4
PolData.inf$MeanAnnualInflation <- c(NA, head(PolData.inf$MeanAnnualInflation, -1))
PolData.inf <- na.omit(PolData.inf)
PolData.infmodel4 <- lm(MeanAnnualInflation ~ PresParty + SenParty - 1, data = PolData.inf)
summary(PolData.infmodel4)
```

```{r}
plot(PolData.infmodel2)
```

Explore the effect of lag on GDP
```{r}
# lag 1
PolData.gdp <- PolData[, c(2, 5, 9, 12, 15)]
PolData.gdp$GDP_AnnualGrowth <- c(NA, head(PolData.gdp$GDP_AnnualGrowth, -1))
PolData.gdp <- na.omit(PolData.gdp)
PolData.gdpmodel1 <- lm(GDP_AnnualGrowth ~ PresParty + TrifectaFlag - 1, data = PolData.gdp)
summary(PolData.gdpmodel1)
```

```{r}
# lag 2
PolData.gdp$GDP_AnnualGrowth <- c(NA, head(PolData.gdp$GDP_AnnualGrowth, -1))
PolData.gdp <- na.omit(PolData.gdp)
PolData.gdpmodel2 <- lm(GDP_AnnualGrowth ~ PresParty + TrifectaFlag - 1, data = PolData.gdp)
summary(PolData.gdpmodel2)
```
Optimum: GDP_AnnualGrowth ~ PresParty and TrifectaFlag with lag 2

```{r}
plot(PolData.gdpmodel2)
```

```{r}
boxplot(GDP_AnnualGrowth~PresParty_label,data=PolData.gdp, xlab="Party", ylab = "GDP Growth" ,col=c("blue", "red"),main ="GDP Annual Growth versus Presidential Party")
boxplot(GDP_AnnualGrowth~TrifectaFlag_label,data=PolData.gdp,xlab="Party", ylab = "GDP Growth",col=c("blue", "grey", "red"),main ="GDP Annual Growth versus TrifectaFlag")
```

```{r}
# lag 3
PolData.gdp$GDP_AnnualGrowth <- c(NA, head(PolData.gdp$GDP_AnnualGrowth, -1))
PolData.gdp <- na.omit(PolData.gdp)
PolData.gdpmodel3 <- lm(GDP_AnnualGrowth ~ PresParty + TrifectaFlag - 1, data = PolData.gdp)
summary(PolData.gdpmodel3)
```

```{r}
# lag 4
PolData.gdp$GDP_AnnualGrowth <- c(NA, head(PolData.gdp$GDP_AnnualGrowth, -1))
PolData.gdp <- na.omit(PolData.gdp)
PolData.gdpmodel4 <- lm(GDP_AnnualGrowth ~ PresParty + TrifectaFlag - 1, data = PolData.gdp)
summary(PolData.gdpmodel4)
```
Since the best performing GDP model is lag 2, it was chosen to perform a residual analysis

```{r}
plot(PolData.gdpmodel2)
```

Explore the effect of lag on Minimum Wage
```{r}
# lag 1
PolData.minWage <- PolData[, c(2, 4, 11, 12, 14)]
PolData.minWage$AdjMinWage <- c(NA, head(PolData.minWage$AdjMinWage, -1))
PolData.minWage <- na.omit(PolData.minWage)
PolData.minWagemodel1 <- lm(AdjMinWage ~ PresParty + HouseParty - 
    1, data = PolData.minWage)
summary(PolData.minWagemodel1)
```

```{r}
# lag 2
PolData.minWage$AdjMinWage <- c(NA, head(PolData.minWage$AdjMinWage, -1))
PolData.minWage <- na.omit(PolData.minWage)
PolData.minWagemodel2 <- lm(AdjMinWage ~ PresParty + HouseParty - 
    1, data = PolData.minWage)
summary(PolData.minWagemodel2)
```

```{r}
# lag 3
PolData.minWage$AdjMinWage <- c(NA, head(PolData.minWage$AdjMinWage, -1))
PolData.minWage <- na.omit(PolData.minWage)
PolData.minWagemodel3 <- lm(AdjMinWage ~ PresParty + HouseParty - 
    1, data = PolData.minWage)
summary(PolData.minWagemodel3)
```
Optimum: AdjMinWage ~ PresParty and HouseParty with lag 3

```{r}
plot(PolData.minWagemodel3)
```

```{r}
boxplot(AdjMinWage~PresParty_label,data=PolData.minWage, xlab="Party", ylab = "Minimum Wage" ,col=c("blue", "red"),main ="Adjusted Minimum Wage versus Presidential Party")
boxplot(AdjMinWage~HouseParty_label,data=PolData.minWage,xlab="Party", ylab = "Minimum Wage",col=c("blue", "red"),main ="Adjusted Minimum Wage versus House Party")
```

```{r }
 plot_minwage<- ggplot(data = PolData, aes(as.numeric(Year), AdjMinWage)) +xlab("Year") + ylab("Minimum Wage") +
  geom_point() +
  ggtitle("Adjusted Minimum Wage over Time") +
  geom_smooth()
plot_minwage
```

```{r}
# lag 4
PolData.minWage$AdjMinWage <- c(NA, head(PolData.minWage$AdjMinWage, -1))
PolData.minWage$AdjMinWage <- c(NA, head(PolData.minWage$AdjMinWage, -1))
PolData.minWage <- na.omit(PolData.minWage)
PolData.minWagemodel4 <- lm(AdjMinWage ~ PresParty + HouseParty - 
    1, data = PolData.minWage)
summary(PolData.minWagemodel4)
```

Explore the effect of lag on Unemployment
```{r}
# lag 1
PolData.unemp <- PolData[, c(2, 4, 8)]
PolData.unemp$UnemploymentRateDecember <- c(NA, head(PolData.unemp$UnemploymentRateDecember, -1))
PolData.unemp <- na.omit(PolData.unemp)
PolData.unempmodel1 <- lm(UnemploymentRateDecember ~ PresParty + HouseParty - 
    1, data = PolData.unemp)
summary(PolData.unempmodel1)
```

```{r}
# lag 2
PolData.unemp$UnemploymentRateDecember <- c(NA, head(PolData.unemp$UnemploymentRateDecember, -1))
PolData.unemp <- na.omit(PolData.unemp)
PolData.unempmodel2 <- lm(UnemploymentRateDecember ~ PresParty + HouseParty - 
    1, data = PolData.unemp)
summary(PolData.unempmodel2)
```

```{r}
# lag 3
PolData.unemp$UnemploymentRateDecember <- c(NA, head(PolData.unemp$UnemploymentRateDecember, -1))
PolData.unemp <- na.omit(PolData.unemp)
PolData.unempmodel3 <- lm(UnemploymentRateDecember ~ PresParty + HouseParty - 
    1, data = PolData.unemp)
summary(PolData.unempmodel3)
```

```{r}
# lag 4
PolData.unemp$UnemploymentRateDecember <- c(NA, head(PolData.unemp$UnemploymentRateDecember, -1))
PolData.unemp <- na.omit(PolData.unemp)
PolData.unempmodel4 <- lm(UnemploymentRateDecember ~ PresParty + HouseParty - 
    1, data = PolData.unemp)
summary(PolData.unempmodel4)
```

```{r}
print("ADJUSTED R-SQUARED VALUES WITH LAGS")
print("no lag     lag 1     lag 2     lag 3     lag 4")
print("inflation")
print(c(summary(inflation.backward)$adj.r.squared, summary(PolData.infmodel1)$adj.r.squared, summary(PolData.infmodel2)$adj.r.squared, summary(PolData.infmodel3)$adj.r.squared, summary(PolData.infmodel3)$adj.r.squared))
print("GDP")
print(c(summary(GDP.backward)$adj.r.squared, summary(PolData.gdpmodel1)$adj.r.squared, summary(PolData.gdpmodel2)$adj.r.squared, summary(PolData.gdpmodel3)$adj.r.squared, summary(PolData.gdpmodel3)$adj.r.squared))
print("minimum wage")
print(c(summary(minimumWage.backward)$adj.r.squared, summary(PolData.minWagemodel1)$adj.r.squared, summary(PolData.minWagemodel2)$adj.r.squared, summary(PolData.minWagemodel3)$adj.r.squared, summary(PolData.minWagemodel4)$adj.r.squared))
print("unemployment")
print(c(summary(unemployment.backward)$adj.r.squared, summary(PolData.unempmodel1)$adj.r.squared, summary(PolData.unempmodel2)$adj.r.squared, summary(PolData.unempmodel3)$adj.r.squared, summary(PolData.unempmodel4)$adj.r.squared))
```

```{r}
plot(PolData.unempmodel2)
```
Examining the qq plot reveals a right skew to the data.  A box cox transformation will be attempted and the response variable transformed.

```{r}
boxcox_unemp=boxcox(unemployment.backward)
lambda=boxcox_unemp$x[which.max(boxcox_unemp$y)]
lambda
```

```{r}
PolData.unempmode.box=lm(((UnemploymentRateDecember^lambda-1)/lambda) ~ PresParty + HouseParty - 
    1, data = PolData.unemp)
summary(PolData.unempmode.box)
```

```{r}
plot(PolData.unempmode.box)
```

Hypothesis Testing

1. Inflation - model: PolData.infmodel2
```{r}
# H0: PresParty = SenParty
a <- c(1, -1)
linearHypothesis(PolData.infmodel2, a, 0)
```
Since the p-value is 0.7317, the null hypothesis can not be rejected at 90%, 95% and 99% (CL > 26.8%). So we can conclude that, at these confidence levels, the effects of the Presidential Party and the Senate Party are not unequal.

2. GDP - model: PolData.gdpmodel2
```{r}
# H0: PresParty = TrifectaFlag
a <- c(1, -1)
linearHypothesis(PolData.gdpmodel2, a, 0)
```
Since the p-value is 0.6265, the null hypothesis can not be rejected at 90%, 95% and 99% (CL > 37.3%). So we can conclude that, at these confidence levels, the effects of the Presidential Party and the Trifecta Flag are not unequal.

3. Minimum Wage - model: minimumWage.backward
```{r}
# H0: PresParty = HouseParty
a <- c(1, -1)
linearHypothesis(PolData.minWagemodel3, a, 0)
```
Since the p-value is 0.08092, the null hypothesis can not be rejected at 95% and 99% (CL > 91.9%). So we can conclude that, at these confidence levels, the effects of the Presidential Party and the House Party are not unequal. The null hypothesis, however, can be rejected at 90% confidence or lower, meaning that we can say that the effect of the two differs with 90% confidence.

4. Unemployment - model: unemployment.backward
```{r}
# H0: PresParty = HouseParty
a <- c(1, -1)
linearHypothesis(unemployment.backward, a, 0)
```
Since the p-value is 0.3899, the null hypothesis can not be rejected at 90%, 95% and 99% (CL > 61%). So we can conclude that, at these confidence levels, the effects of the Presidential Party and the House Party are not unequal.
